---
title: "Antibiotic"
author: "Bashir Ssuna"
date: "2023-10-28"
output: html_document
---


```{r, echo=T, warning=FALSE}
library(readxl)
abx <- read_excel("D:/ACADEMIA/Research/Antibiotics/new/outcomes2.xls")

colnames(abx)


```

```{r}
library(tidyverse)
library(viridis)
library(hrbrthemes)
library(scales)

ggplot(abx, aes(x = age)) +
  geom_histogram(binwidth = 1, fill = "red", color = "black", alpha = 0.7) +
  theme_minimal() +
  labs(title = "Distribution of Participants' age", x = "Participants' age", y = "Number of Participants")

```

```{r}
str(abx$age)
str(abx$study_arm)
str(abx$Primary_outcome)
```

```{r}
# Labeling outcome variables
abx$Primary_outcome <- factor(abx$Primary_outcome, levels = c(0, 1), labels = c("No Infection", "Infection"))

abx$Secondary_outcome <- factor(abx$Secondary_outcome, levels = c(0, 1), labels = c("No Infection", "Infection"))
```


Lets look at the distribution of age vs study arm and primary outcome


```{r, warning=FALSE}
plot_1 = abx %>%
  ggplot( aes(x= study_arm, y=age, fill= Primary_outcome )) +
    geom_boxplot(aes( color = study_arm)) +scale_fill_viridis(discrete = TRUE, alpha=0.4) +
    geom_jitter(color="blue", size=0.5, alpha=0.5) +
    theme_ipsum() +
    theme(
      legend.position="right",
      plot.title = element_text(size=11), axis.text.x = element_text(size = 8)
    ) +
    ggtitle("Distribution of Participants' age according to Study arm and Primary Outcome") +
    xlab("Study arm ")+ylab("Age of participants")+labs(fill = "Primary Outcome", color="Study arm")+scale_x_discrete(guide = guide_axis(n.dodge=1))

print(plot_1)

```

Visualizing: Proportion of primary outcomes in each study arm

```{r}
# Compute the counts for each group combination
counts <- as.data.frame(table(abx$Primary_outcome, abx$study_arm))

# Calculate the total counts for each study_arm
study_arm_totals <- table(abx$study_arm)

# Calculate the proportions for each Primary_outcome within each study_arm
counts$proportion <- counts$Freq / study_arm_totals[as.character(counts$Var2)]

# Convert proportions to percentages and round off to 1 decimal place
counts$percentage <- round(counts$proportion * 100, 1)

# Convert the percentage to a label with "%" sign
counts$percentage_label <- paste0(counts$percentage, "%")

# Rename columns for clarity
colnames(counts) <- c("Primary_outcome", "study_arm", "Freq", "proportion", "percentage", "percentage_label")

# Plot using ggplot
Plot4<- ggplot(counts, aes(x = Primary_outcome, y = Freq, fill = study_arm)) + 
  geom_bar(stat = "identity", position = "dodge") + 
  geom_text(aes(label = percentage_label, y = Freq/2), position = position_dodge(width = 0.9), vjust= -2.5) +
  labs(title = "Proportion Participants accoring to Study arm and Primary outcome", 
       x = "Primary outcome", 
       y = "Proportion of patients", 
       fill = "Study arm") +
  theme_minimal()

print(Plot4)
```

```{r}

# Compute the counts for each group combination
counts_sec <- as.data.frame(table(abx$Secondary_outcome, abx$study_arm))

# Calculate the total counts for each study_arm
study_arm_totals_sec <- table(abx$study_arm)

# Calculate the proportions for each Secondary_outcome within each study_arm
counts_sec$proportion <- counts_sec$Freq / study_arm_totals_sec[as.character(counts_sec$Var2)]

# Convert proportions to percentages and round off to 1 decimal place
counts_sec$percentage <- round(counts_sec$proportion * 100, 1)

# Convert the percentage to a label with "%" sign
counts_sec$percentage_label <- paste0(counts_sec$percentage, "%")

# Rename columns for clarity
colnames(counts_sec) <- c("Secondary_outcome", "study_arm", "Freq", "proportion", "percentage", "percentage_label")


# Plot using ggplot
plot2 <- ggplot(counts_sec, aes(x = Secondary_outcome, y = Freq, fill = study_arm)) + 
  geom_bar(stat = "identity", position = "dodge") + 
  geom_text(aes(label = percentage_label, y = Freq/2), position = position_dodge(width = 0.9), vjust= -2.5) +
  labs(title = "Proportion Participants accoring to Study arm and Secondary outcome", 
       x = "Secondary outcome", 
       y = "Proportion of patients", 
       fill = "Study arm") +
  theme_minimal()

print(plot2)
```

Lets create a dataframe for Outcomes and Followup periods to plot the proportions of outcomes per day

```{r}

# Create dataframe in R

# Data
follow_up_periods <- c("Day 1", "Day 2", "Day 3", "Day 4", "Day 5", "Day 6")
total_participants <- c(260, 256, 247, 244, 244, 256)
primary_outcome <- c(7, 7, 10, 6, 12, 15)
secondary_outcome <- c(18, 12, 14, 11, 13, 16)

# Create dataframe
df <- data.frame(Follow_up_period = follow_up_periods, 
                 Total_Participants = total_participants, 
                 Primary_Outcome = primary_outcome, 
                 Secondary_Outcome = secondary_outcome)

# View dataframe
print(df)


```


```{r}
# Compute row with totals
totals <- data.frame(Follow_up_period = "Total",
                    Total_Participants = sum(df$Total_Participants),
                    Primary_Outcome = sum(df$Primary_Outcome),
                    Secondary_Outcome = sum(df$Secondary_Outcome))

# Append the totals row to the original dataframe
df_with_totals <- rbind(df, totals)

# Display the dataframe with the totals row
df_with_totals

```


```{r}
# Compute percentages for primary and secondary outcomes and round to two decimal places
df_with_totals$Primary_Percentage <- round((df_with_totals$Primary_Outcome / df_with_totals$Total_Participants) * 100, 2)

df_with_totals$Secondary_Percentage <- round((df_with_totals$Secondary_Outcome / df_with_totals$Total_Participants) * 100, 2)

df_with_totals
df2 <- df_with_totals

```


```{r}
df2
colnames(df2)
str(df2)

```


Plotting proportions of outcomes
N.B: This is not the same as incidence


```{r}

Plot3<- ggplot(df2, aes(x = as.factor(Follow_up_period))) +
  geom_line(aes(y = Primary_Percentage, color = "Primary Outcome", group = 1), size = 1) +
  geom_line(aes(y = Secondary_Percentage, color = "Secondary Outcome", group = 2), size = 1, linetype = "dashed") +
  
  # Adding dots for each data point
  geom_point(aes(y = Primary_Percentage, color = "Primary Outcome"), size = 3) +
  geom_point(aes(y = Secondary_Percentage, color = "Secondary Outcome"), size = 3) +
  
  # Adding data labels for Primary_Percentage and Secondary_Percentage with "%" sign
  geom_text(aes(y = Primary_Percentage, label = paste0(Primary_Percentage, "%")), vjust = 1.5, hjust = 0.5, size = 3) +
  geom_text(aes(y = Secondary_Percentage, label = paste0(Secondary_Percentage, "%")), vjust = -0.5, hjust = 0.5, size = 3) +
  
  # Adjusting the y-axis limits
  ylim(0, 7) +
  
  labs(title = "Study Outcomes over Follow-up Periods",
       y = "Percentage of Participants", 
       x = "Follow-up Period",
       color = "Outcome") +
  theme_minimal()

print(Plot3)


```
Now lets look at the entire dataset

```{r}
abx2 <- read_excel("D:/ACADEMIA/Research/Antibiotics/All_data.xls")

colnames(abx2)
```

Chi square for a crude association between exposure and primary outcome

```{r}
tab2 <- table(abx$study_arm2, abx$Primary_outcome)
tab2
chi<-chisq.test(tab2)
fisher <- fisher.test(tab2)
chi
fisher

```

Chi square for a crude association between exposure and secondary outcome

```{r}
tab2 <- table(abx$study_arm2, abx$Secondary_outcome)
tab2
chi<-chisq.test(tab2)
fisher <- fisher.test(tab2)
chi
fisher


```

Calculation of Incidence Rate Ratio of the primary outcome
Incidence = (Total number of events)/(Total person time)
Person Time = Number of followed-up individuals X days of follow-up

```{r}
# Calculate person-time for each arm
person_time_intervention <-  c(130*1)+ (129*2)+ (117*3)+ (121*4)+ (119*5)+ (126*6)
person_time_control <- c(130*1)+ (127*2)+ (130*3)+ (123*4)+ (125*5)+ (130*6)


# Given data
outcomes_intervention <- 14
outcomes_control <- 20

# Calculate incidence rates for each arm
IR_control = outcomes_control/person_time_control
IR_intervention = outcomes_intervention / person_time_intervention


# Calculate incidence rate ratio
IRR = IR_intervention / IR_control


# Load necessary library
#install.packages("epitools")
#library(epitools)

# Calculate 95% confidence intervals and p-value for IRR
# Using irr function from epitools package
#result = riskratio.wald(c(outcomes_intervention, outcomes_control), c(person_time_intervention, person_time_control))

#list(IR_intervention = IR_intervention, IR_control = IR_control, IRR = IRR, CI = result$measure, p_value = result$p.value)

Total_person_time = person_time_control+person_time_intervention

person_time_intervention
person_time_control
Total_person_time

IR_intervention*1000
IR_control*1000
IRR

```

95% CI for IRR

```{r}
# Given data
outcomes_intervention <- 14
outcomes_control <- 20
person_time_intervention <-  c(130*1)+ (129*2)+ (117*3)+ (121*4)+ (119*5)+ (126*6)
person_time_control <- c(130*1)+ (127*2)+ (130*3)+ (123*4)+ (125*5)+ (130*6)
number_control = 130
number_intervention = 130
# Conduct the test
test_result <- prop.test(x = c(outcomes_intervention, outcomes_control),
                         n = c(number_intervention, number_control))

# Extract the p-value and confidence interval
proportion_difference <- (outcomes_intervention/130)*100 - (outcomes_control/130)*100

p_value <- test_result$p.value
conf_int <- (test_result$conf.int)*100

# Print the results
proportion_difference
print(paste("P-value:", p_value))
print(paste("95% CI for the difference between proportions:", conf_int[1], "to", conf_int[2]))


```



```{r}
# Install and load the epiR package if you haven't already
#if (!require(epiR)) {
  
    #install.packages("epiR")
    #library(epiR)
#}

# load data
Exposed_postive <- 20
Exposed_negative <- 14
Outcome_positive <- 130
Outcome_negative <- 130

# Calculate the incidence rate ratio (IRR) and its 95% confidence interval
result <- epiR::epi.2by2(dat = matrix(c(outcomes_intervention, Outcome_negative - Exposed_negative,
                                        outcomes_control, Outcome_positive - Exposed_postive),
                                      nrow = 2, byrow = TRUE),
                         method = "cohort.count")

# Print the IRR and 95% CI
print(result)


```



```{r}
incidence_rate_intervention <- 5.44 / 1000 # per person-day
incidence_rate_control <- 7.49 / 1000 # per person-day
person_time_intervention <- 2574 # person-days
person_time_control <- 2671 # person-days

```



```{r}
# Calculate 95% confidence intervals for incidence rates
ci_low_control = IR_control - 1.96 * sqrt(IR_control / person_time_control)
ci_high_control = IR_control + 1.96 * sqrt(IR_control / person_time_control)

ci_low_intervention = IR_intervention - 1.96 * sqrt(IR_intervention / person_time_intervention)
ci_high_intervention = IR_intervention + 1.96 * sqrt(IR_intervention / person_time_intervention)

list(
  ci_control = c(ci_low_control, ci_high_control)*1000,
  ci_intervention = c(ci_low_intervention, ci_high_intervention)*1000
)


```


```{r}
# Given data
outcomes_intervention <- 22
outcomes_control <- 24
person_time_intervention <-  c(130*1)+ (129*2)+ (117*3)+ (121*4)+ (119*5)+ (126*6)
person_time_control <- c(130*1)+ (127*2)+ (130*3)+ (123*4)+ (125*5)+ (130*6)

# Create the 2x2 table
table_data <- data.frame(
  Study_Arm = c("Intervention", "Control"),
  Number_of_Outcomes = c(outcomes_intervention, outcomes_control),
  Person_time = c(person_time_intervention, person_time_control)
)

print(table_data)


```


Lets now analyze the data with outcomes only

```{r}
# Read the data from the CSV file
outcomes <- read_excel("D:/ACADEMIA/Research/Antibiotics/outcomes.xlsx")
colnames(outcomes)

# Convert data from wide to long format for ggplot
outcomes_long <- gather(outcomes, key = "Outcome", value = "Percentage", -`Follow-up period`)

outcomes_long
# Plot the line graph using ggplot2
ggplot(outcomes_long, aes(x = `Follow-up period`, y = Percentage, color = Outcome, group = Outcome)) +
  geom_line() +
  geom_point() +geom_text(aes(label = Percentage), 
            position = position_dodge(width = 0.9), 
            vjust = -2.0, 
            color = "black", 
            size = 3.5) +
  theme_minimal() +
  labs(title = "Outcomes by Follow-up Period",
       x = "Follow-up Period",
       y = "Percentage") +
  scale_color_brewer(palette = "Set1")  # This will use a color palette suitable for distinct colors


```


```{r}
# Plot the bar graph using ggplot2
ggplot(outcomes_long, aes(x = `Follow-up period`, y = Percentage, fill = Outcome)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_minimal() +
  labs(title = "Outcomes by Follow-up Period",
       x = "Follow-up Period",
       y = "Percentage") +
  scale_fill_brewer(palette = "Set1") 

```


```{r}
# Define the width for dodging
dodge_width <- 0.7

# Plot the bar graph using ggplot2
ggplot(outcomes_long, aes(x = `Follow-up period`, y = Percentage, fill = Outcome)) +
  geom_bar(stat = "identity", position = position_dodge(width = dodge_width), width = dodge_width)+ #+
 # geom_text(aes(label = Percentage), 
  #          position = position_dodge(width = dodge_width), 
   #         vjust = -2.0, 
    #        color = "black", 
     #       size = 2.5) +
  theme_minimal() +
  labs(title = "Outcomes by Follow-up Period",
       x = "Follow-up Period",
       y = "Percentage") +
  scale_fill_brewer(palette = "Set1")  # This will use a color palette suitable for distinct colors

```


```{r}
# Create dataframe in R

# Data
follow_up_periods <- c("Day 1", "Day 2", "Day 3", "Day 4", "Day 5", "Day 6")
total_participants <- c(260, 256*2, 247*3, 244*4, 244*5, 256*6)
primary_outcome <- c(7, 7, 10, 6, 12, 15)
secondary_outcome <- c(18, 12, 14, 11, 13, 16)

# Create dataframe
df <- data.frame(Follow_up_period = follow_up_periods, 
                 Total_Participants = total_participants, 
                 Primary_Outcome = primary_outcome, 
                 Secondary_Outcome = secondary_outcome)

# View dataframe
print(df)

```



```{r}

# Compute row with totals
totals <- data.frame(Follow_up_period = "Total",
                    Total_Participants = sum(df$Total_Participants),
                    Primary_Outcome = sum(df$Primary_Outcome),
                    Secondary_Outcome = sum(df$Secondary_Outcome))

# Append the totals row to the original dataframe
df_with_totals <- rbind(df, totals)

# Display the dataframe with the totals row
df_with_totals

```


```{r}
# Compute percentages for primary and secondary outcomes and round to two decimal places
df_with_totals$Primary_Percentage <- round((df_with_totals$Primary_Outcome / df_with_totals$Total_Participants) * 100, 2)

df_with_totals$Secondary_Percentage <- round((df_with_totals$Secondary_Outcome / df_with_totals$Total_Participants) * 100, 2)

df_with_totals
df2 <- df_with_totals

```


```{r}

# Compute percentages for primary and secondary outcomes and round to two decimal places
df$Primary_Percentage <- round((df$Primary_Outcome / df$Total_Participants) * 100, 2)

df$Secondary_Percentage <- round((df$Secondary_Outcome / df$Total_Participants) * 100, 2)


```

Plot of Incidence rates of primary outcome for every follow-up period in each study arm

```{r}
ggplot(df, aes(x = as.factor(Follow_up_period))) +
  geom_line(aes(y = Primary_Percentage, color = "Primary Outcome", group = 1), size = 1) +
  geom_line(aes(y = Secondary_Percentage, color = "Secondary Outcome", group = 2), size = 1, linetype = "dashed") +
  
  # Adding dots for each data point
  geom_point(aes(y = Primary_Percentage, color = "Primary Outcome"), size = 2) +
  geom_point(aes(y = Secondary_Percentage, color = "Secondary Outcome"), size = 2) +
  
  # Adding data labels for Primary_Percentage and Secondary_Percentage with "%" sign
  geom_text(aes(y = Primary_Percentage, label = paste0(Primary_Percentage, "%")), vjust = 1.5, hjust = 0.5, size = 3) +
  geom_text(aes(y = Secondary_Percentage, label = paste0(Secondary_Percentage, "%")), vjust = -0.5, hjust = 0.5, size = 3) +
  
  # Adjusting the y-axis limits
  ylim(0, 7) +
  
  labs(title = "Incidence of Study Outcomes over Follow-up Periods",
       y = "Incidence (%)", 
       x = "Follow-up Period",
       color = "Outcome") +
  theme_minimal()


```

Lets combine primary and secondary outcomes per study arm on one graph

```{r}
# Create dataframe in R for incidence per study_arm

# Data
follow_up_periods <- c("Day 1", "Day 2", "Day 3", "Day 4", "Day 5", "Day 6")
No_antibiotics <- c(130, 127*2, 130*3, 123*4, 125*5, 130*6)
Antibiotics <- c(130, 129*2, 117*3, 121*4, 119*5, 126*6)
primary_outcome <- c(3, 3, 7, 2, 9, 11)
secondary_outcome <- c(4, 4, 3, 4, 3, 4)

# Create dataframe
df <- data.frame(Follow_up_period = follow_up_periods, 
                 No_antibiotics = No_antibiotics,
                 Antibiotics =Antibiotics,
                 Primary_Outcome = primary_outcome, 
                 Secondary_Outcome = secondary_outcome)

# View dataframe
print(df)

```


```{r}
# Compute percentages for primary and secondary outcomes and round to two decimal places in Control
df$Primary_Percentage_noabx <- round((df$Primary_Outcome / df$No_antibiotics) * 100, 2)

df$Secondary_Percentage_noabx <- round((df$Secondary_Outcome / df$No_antibiotics) * 100, 2)

# Compute percentages for primary and secondary outcomes and round to two decimal placesin Intervention
df$Primary_Percentage_abx <- round((df$Primary_Outcome / df$Antibiotics) * 100, 2)

df$Secondary_Percentage_abx <- round((df$Secondary_Outcome / df$Antibiotics) * 100, 2)

df
```
```{r}
# Convert data from wide to long format for ggplot
#df_long <- gather(Study_arm, key = "Study_arm", value = "Person_time", -`Follow-up period`)

# Gather the variables into Study_arm
gathered_data <- gather(df, key = "Study_arm", value = "Person_time", No_antibiotics, Antibiotics)
gathered_data

```


```{r}
ggplot(df, aes(x = as.factor(Follow_up_period))) +
  geom_line(aes(y = Primary_Percentage_abx, color = "Primary Outcome antibiotics arm", group = 1), size = 1) +
  geom_line(aes(y = Primary_Percentage_noabx, color = "Primary Outcome No antibiotics arm", group = 2), size = 1, linetype = "dashed") +
  
  # Adding dots for each data point
  geom_point(aes(y = Primary_Percentage_abx, color = "Primary Outcome antibiotics arm"), size = 2) +
  geom_point(aes(y = Primary_Percentage_noabx, color = "Primary Outcome No antibiotics arm"), size = 2) +
   geom_line(aes(y = Secondary_Percentage_abx, color = "Secondary Outcome antibiotics arm", group = 1), size = 1) +
  geom_line(aes(y = Secondary_Percentage_noabx, color = "Secondary Outcome No antibiotics arm", group = 2), size = 1, linetype = "dashed") +
  
  # Adding dots for each data point
  geom_point(aes(y = Secondary_Percentage_abx, color = "Secondary Outcome antibiotics arm"), size = 2) +
  geom_point(aes(y = Secondary_Percentage_noabx, color = "Secondary Outcome No antibiotics arm"), size = 2) +
  
  # Adding data labels for Primary_Percentage and Secondary_Percentage with "%" sign
  #geom_text(aes(y = Primary_Percentage_abx, label = paste0(Primary_Percentage_abx, "%")), vjust = -0.5, hjust = 0.5, size = 3) +
  #geom_text(aes(y = Primary_Percentage_noabx, label = paste0(Primary_Percentage_noabx, "%")), vjust = -0.5, hjust = 0.5, size = 3) +
  
  # Adjusting the y-axis limits
  ylim(0, 7) +
  
  labs(title = "Incidence of Outcomes over Follow-up Periods in study arms",
       y = "Incidence (%)", 
       x = "Follow-up Period",
       color = "Study arm") +
  theme_minimal()



```

